Appendix B: System size expansion of the master equation for demographic noise

### 2017-10-31 appendixB/appendixB.Rmd
WIP

---
title: "Appendix B: System size expansion of the master equation for demographic noise"
author: 
  name: "Carl Boettiger"
  affiliation: a
address:
  - code: a
    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
#layout: 5p  
bibliography: canonical.bib
output: 
  rticles::elsevier_article:
    includes:
      in_header: header.tex

---

<!--
## The Poisson process description

The Poisson process is a common description of demographic noise.  It arises as the limit of a binomial process (coin flipping) when the events are rare and the population large.  This model is commonly applied as follows.  Consider a model of the form:

\begin{align}
N_{t+1} = f(N_t)
\end{align}

This approach chooses with probability $P$ the number in the next generation $N_{t+1}$ at random from a Poisson distribution with mean $f(N)$:

\begin{align}
P(N_t)_{t+1} = \text{Poisson}(N, f(N_t)), 
\end{align}

while the variance of the Poisson distribution will also be $f(N)$.  If the population has some equilibrium size $N^* \neq 0$ such that $f(N^*) = N^*$, then at equilibrium the population will have size $N^*$ and fluctuate around that value with standard deviation $\sqrt{N^*}$.  In the continuous time limit this approaches a Gaussian with the same mean and variance.  Observe that $P(0) > 0$ for any $N$, hence a population can always go extinct.  Further, if we naturally assume the population is isolated, then $f(0) = 0$, and extinction is the only long-term stable state.  This description provides a first pass at describing demographic noise, and captures some of its important phenomena:

- Finite populations fluctuate around their quasi-equilibrium.
- Any isolated population that cannot grow without bound will eventually go extinct.
-  The magnitude of the steady-state fluctuations relative to the equilibrium decreases with increasing population size (as $\sqrt{N^*}/N^*$), decreasing the extinction risk of larger populations proportionally.

While we have started to capture one aspect of demographic noise (populations are finite), we have ignored the second (population dynamics come from birth and death processes).  Too often we stop at this description of demographic noise.  By using only $f(N)$, we have not specified births and deaths separately.  This is quite common in population models, where the choice of $f(N)$ does not obviously suggest which are the birth and which are the death terms.  Our second introductory example illustrates the importance of doing so.  
-->

@@

\begin{figure}
\begin{center}
\includegraphics[width=.25\textwidth]{markov}
\caption{The birth-death process as a Markov process.}\label{markov}
\end{center}
\end{figure}

@@

@@

\begin{align}
\frac{\ud }{\ud t} \langle n \rangle &= \sum n (\mathbb E^{-1} - 1) b_n p_n + \sum n (\mathbb E -1) d_n p_n \nonumber \\ 
&= \sum  b_n p_n(\mathbb E - 1)n  + \sum d_n p_n(\mathbb E^{-1} -1) n \nonumber \\
&= - \langle d_n \rangle +  \langle b_n \rangle \label{mean} \\
\intertext{and similarly}
\frac{\ud }{\ud t} \langle n^2 \rangle &= 2 \langle n (b_n - d_n) \rangle + \langle b_n+d_n \rangle \label{var}
\end{align}

@@

Where we have the part of the birth and death functions that depend on the macroscopic variable $\phi$ alone.  This is known as the macroscopic equation, and corresponds to the density equations commonly written down.  We call this difference $\alpha_1(\phi)$ as a shorthand\footnote{This notation suggests it is the first jump moment, defined as first moment of the transition rate between states in a Markov process.  It so happens this term will give the macroscopic law for any Markov process, not only a birth-death process.  The second moment of the transition rates, $a_2$, will be for us simply the sum of the birth and death rates.  We use this notation as it generalizes to processes that have a different master equation from \eqref{master1}.}

@@

Where we define $\alpha_2(\phi) = b(\phi)+d(\phi)$ as an analogous short hand.  From this it is a straight forward exercise to calculate the moments of the distribution (multiply by $\xi$, $\xi^2$ and integrate),

@@

@@

\begin{align}
\dot n = c n \left(1 - \frac{n}{N}\right) - e n \label{levins}
\end{align}

@@

# Alternate derivation: the diffusion approximation

>>>>

---
title: "Appendix B: System size expansion of the master equation for demographic noise"
author: 
  name: "Carl Boettiger"
  affiliation: a
address:
  - code: a
    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
layout: 5p  
bibliography: canonical.bib
output: rticles::elsevier_article
preamble: "\\newcommand{\\ud}{\\mathrm{d}}"
---

@@

@@

\begin{figure}
\begin{center}
\includegraphics[width=.25\textwidth]{markov}
\caption{Birth and death as state-dependent transition rates in a Markov process.}\label{markov}
\end{center}
\end{figure}

@@

\begin{align}
\frac{\ud }{\ud t} \langle n \rangle &= \sum n (\mathbb E^{-1} - 1) b_n p_n + \sum n (\mathbb E -1) d_n p_n \nonumber \\ 
&= \sum  b_n p_n(\mathbb E - 1)n  + \sum d_n p_n(\mathbb E^{-1} -1) n \nonumber \\
&= - \langle d_n \rangle +  \langle b_n \rangle \label{mean} \\
\intertext{and similarly the second moment is}
\frac{\ud }{\ud t} \langle n^2 \rangle &= 2 \langle n (b_n - d_n) \rangle + \langle b_n+d_n \rangle \label{var}
\end{align}

@@

Where we have the part of the birth and death functions that depend on the macroscopic variable $\phi$ alone.  This is known as the macroscopic equation, and corresponds to the density equations commonly written down.  Following @vanKampen, we will call this difference $\alpha_1(\phi)$ as a shorthand\footnote{This notation suggests it is the first jump moment, defined as first moment of the transition rate between states in a Markov process.  It so happens this term will give the macroscopic law for any Markov process, not only a birth-death process.  The second moment of the transition rates, $a_2$, will be for us simply the sum of the birth and death rates.  We use this notation as it generalizes to processes that have a different master equation from \eqref{master1}.}

@@

Where we define $\alpha_2(\phi) = b(\phi)+d(\phi)$ as an analogously.  From this it is a straight forward exercise to calculate the moments of the distribution (multiply by $\xi$, $\xi^2$ and integrate),

@@

## Stochastic Differential Equations

The partial differential equation (PDE) given by \eqref{FP} is sometimes referred to as the Fokker-Planck equation. It is possible to prove this PDE is equivalent to as the stochastic differntial equation (SDE),

\begin{equation}
\textrm{d} X_t = \mu(X_t) \textrm{d}t + \sigma(X_t) \textrm{d}B_t \label{SDE}
\end{equation}

which, despite the name, it is important to remember that this is actually an integral equation in the Itô calculus, and not a derivative. Note the separation of the differential elements, which are meant as shorthand for implicit integrals:

\begin{equation}
X_{t+s} - X{t}  = \int_t^{t+s} \mu(X_{\tau}, \tau) \textrm{d}\tau + \int_t^{t+s} \sigma(X_{\tau}, \tau) \textrm{d}B\tau
\end{equation}

The first integral is the ordinary Lebesgue integral, while the second is an Itô integral, which integrates over the stochastic process $B_t$ (Brownian motion, also termed a Weiner process). The Itô calculus follows different rules from the classic calculus, largely due to the fact that an Itô integral has non-zero quadratic variation. (Crudely put, $\textrm{d}x^2 \sim 0$, but $\textrm{d}B^2 \sim dx$).  Though it is common to formulate stochastic models directly as SDEs, it is worth remembering that the SDE is the result of a small noise approximation in the system size expansion, and can break down under larger deviations, even when the mean-field equation remains approximately accurate.  @Ovaskainen2010 provides an overview literature in that addresses this limit and discusses the use of the WKB^[Wentzel, Kramers & Brillouin. Naturally this is KWB in Holland, BWK in France, and JWKB in England (for Jeffreys). Note that the physicist Gregor Wentzel is not the same as the mathematician Alexander Wentzell of the Fredilin-Wentzell theorem for large deviations years later, though I believe the former's result can be derived directly from the latter's.] method to address this case.

@@

\begin{align}
\frac{\ud n}{\ud t} = c n \left(1 - \frac{n}{N}\right) - e n \label{levins}
\end{align}

@@

<!-- Include environmental noise example via stochastic K? -->

# An alternate derivation: the diffusion approximation
